---
title: "DSC_1107_FA2"
author: "Frances Aneth Rosales"
date: "`r Sys.Date()`"
output:
  html_document:
    css: FA2.css
    code_folding: hide
---
<style>
  body {
    text-align: justify;
  }
</style>


<br>
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<br><br>

<h3> Applying the Packages First </h3>

```{r}
library(tidyverse) 
library(ggrepel)
library(kableExtra) 
library(cowplot)
```

<br><br><br>
<b>

# 1 Wrangle (35 points for correctness; 5 points for presentation)

</b>

<h1> 1.1 Import (5 points) </h1>
<h3>
<li>  Import the data into a tibble called mlb_raw and print it. </li>
 </h3>

```{r files import}
load("ml_pay.rdata")
library(tibble)
mlb_raw <- as_tibble(ml_pay)
print(mlb_raw)
View(mlb_raw)
```


<br><br>
<h3>
<li>  How many rows and columns does the data have?  </li>
</h3>

```{r rowcol}
mlb_rows <- nrow(mlb_raw)
mlb_columns <- ncol(mlb_raw)

cat("Number of rows of ML Pay:", mlb_rows, "\n")
cat("Number of columns of ML Pay:", mlb_columns, "\n")

```

<br><br>
<h3>
<li> Does this match up with the data description given above?</li>
</h3>
<h3> 

In accordance of the data that I imported into the table with the given  data description, it is indeed similar data as assumed.

Therefore, we can now analyze the data of Major League Baseball (MLB) teams' payroll.

</h3>

<br><br>

<h1> 1.2 Tidy (15 points) </h1>

<h4> 
We need to change the variables containing the following <br>
mlb_aggregate c: aggregate data<br>
mlb_yearly: year-by-year data<br>
mlb_total: columns named team,payroll_aggregate, pct_wins_aggregate<br>
mlb_yearly: contain columns named team, year, payroll, pct_wins, num_wins<br>

</h4>

<h5> mlb_aggregate tibble </h5>


```{r the_mlb_aggregate}
library(tidyverse)
mlb_raw <- mlb_raw %>%
  rename_all(str_to_lower) %>%  
  rename(team = team.name.2014, avgwin = avgwin)  

mlb_raw_aggre <- mlb_raw %>%
  rename_all(str_to_lower) %>% 
  rename( the_payroll = payroll )



mlb_aggregate <- mlb_raw_aggre %>%
  select(team, starts_with("the_payroll"), matches("^X.*\\.pct$")) %>%
  rename_with(~ "payroll_aggregate", starts_with("the_payroll")) %>%
  mutate(pct_wins_aggregate = rowMeans(select(., matches("^X.*\\.pct$")))) %>%
  select(team, starts_with("payroll_aggregate"), pct_wins_aggregate)
print(mlb_aggregate)


```


<br><br>
<b>

<h3> mlb_yearly tibble </h3>

</b>

```{r the_mlb_yearly}

mlb_raw_yearly <- mlb_raw %>%
  rename_all(str_to_lower) %>%  
  rename_with(~ paste0("pct_wins_", str_remove(., "^X")), matches("^X.*\\.pct$")) %>%
  rename(the_payroll = payroll)
  
mlb_yearly <- mlb_raw_yearly %>%
  select(team, the_payroll, starts_with("p"), starts_with("x"), starts_with("wins_aggregate")) %>%
  rename_with(
    ~ str_replace(., "^p(\\d+)$", "payroll_\\1"),
    starts_with("p")
  ) %>%
  rename_with(
    ~ str_replace(., "^x(\\d+)$", "num_wins_\\1"),
    starts_with("x")
  ) %>%
  rename_with(
    ~ str_replace(., "^wins_aggregate", "wins_aggregate"),
    starts_with("wins_aggregate")
  )
print(mlb_yearly)

```
<br>
```{r rowcossl}
mlb_aggregaterows <- nrow(mlb_aggregate)
mlb_yearlyrows <- nrow(mlb_yearly)

cat("Number of rows of mlb_aggregate:", mlb_aggregaterows, "\n")
cat("Number of rows of mlb_yearly  :", mlb_yearlyrows, "\n")

```
<br><br>


<h1> 1.3 Quality control (15 points) </h1>
<b>
<h3> mlb_aggregate_computed tibble </h3>
</b>


```{r the_mlb_aggregate_computed}

library(tidyverse)
mlb_aggregate_computed <- mlb_yearly %>%
  group_by(team) %>%
  summarise(
    payroll_aggregate_computed = sum(across(starts_with("payroll_")), na.rm = TRUE),  # Total team payroll
    pct_wins_aggregate_computed = mean(sum(across(starts_with("pct_wins_x")), na.rm = TRUE), na.rm = TRUE) / 17  ) 

mlb_aggregate_joined <- mlb_aggregate %>%
  left_join(mlb_aggregate_computed, by = "team")
print(mlb_aggregate_joined)



```

<br><br>

<h1> <b> Create Scatter Plots </b> </h1> 
```{r scatter plots}

mlb_aggregate_joined$team <- as.character(mlb_aggregate_joined$team)

library(gridExtra)

library(ggplot2)

plot_payroll <- ggplot(mlb_aggregate_joined, aes(x = payroll_aggregate, y = payroll_aggregate_computed)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  labs(title = "Scatter Plot of Payroll Aggregate",
       x = "Payroll Aggregate",
       y = "Computed Payroll Aggregate")


plot_pct_wins <- ggplot(mlb_aggregate_joined, aes(x = pct_wins_aggregate, y = pct_wins_aggregate_computed)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "blue") +
  labs(title = "Scatter Plot of Win Percentage Aggregate",
       x = "Win Percentage Aggregate",
       y = "Computed Win Percentage Aggregate")
library(grid)

common_title <- textGrob("Figure 1.1", gp = gpar(fontsize = 12, fontface = "bold"))

grid.arrange(
  arrangeGrob(plot_payroll, plot_pct_wins, ncol = 2),
  common_title,
  heights = c(4, 0.5)  # Adjust the heights as needed
)

```
<h4> Using function ggplot, getting the plot of <b> payroll_aggregate_computed versus payroll_aggregate </b> and <b> pct_wins_aggregate_computed versus pct_wins_aggregate </b>, we have concluded the plotting shows a linear model which pertains into a proportion increment of data. This will also that the data is close to similar to each other as desired.</h4>



<br><br><br>
<b>

# 2 Explore (50 points for correctness; 10 points for presentation) 

</b>
<h4>Now that the data are in tidy format, we can explore them by producing visualizations and summary statistics.<h4>

<h1> 2.1 Payroll across years(15 points) </h1>

<h3>
<li>  Plot payroll as a function of year for each of the 30 teams, faceting the plot by team and adding a red dashed horizontal line for the mean payroll across years of each team. </li>
 </h3>


```{r payroll_plot}
library(ggplot2)

the_payroll_plot_now <-mlb_yearly %>%
  gather(key = "year", value = "payroll", starts_with("payroll_")) %>%
  ggplot(aes(x = as.numeric(sub("payroll_", "", year)), y = payroll)) +
  geom_point() +
  facet_wrap(~ team, scales = "free_y") +
  geom_hline(aes(yintercept = mean(payroll, na.rm = TRUE)), linetype = "dashed", color = "red") +
  labs(title = "Payroll as a Function of Year for Each Team",
       x = "Year",
       y = "Payroll") +
  theme_minimal()


library(grid)

payroll_plot_title <- textGrob("Figure 1.2", gp = gpar(fontsize = 12, fontface = "bold"))

grid.arrange( the_payroll_plot_now,
  payroll_plot_title,
  heights = c(4, 0.5) 
)

```

<h4> Using function ggplot and gather, getting the plot of payroll as a function of year for each of the 30 teams. We can now analyze together the plot of payroll of each team over the years of 1998 to 2014. 
We can also say that the mean of the payroll of each team are not equal together as the red horizontal line varied for each team.</h4>

<br><br>


<h3>
<li>  Using dplyr, identify the three teams with the greatest payroll_aggregate_computed, and print a table of these teams and their payroll_aggregate_computed. </li>
 </h3>

```{r dplyr_greteast_payroll_aggregate_computed}

library(dplyr)


top_teams_table <- mlb_aggregate_joined %>%
  top_n(3, payroll_aggregate_computed) %>%
  select(team, payroll_aggregate_computed)

print(top_teams_table)
```

<br><br>


<h3>
<li>  Using dplyr, identify the three teams with the greatest percentage increase in payroll from 1998 to 2014 (call it pct_increase), and print a table of these teams along with pct_increase as well as their payroll figures from
1998 and 2014.
  </li>
 </h3>

```{r pct_increase_year}
library(dplyr)

mlb_yearly_increase <- mlb_yearly %>%
  select(team, matches("^payroll_")) %>%
  rename_with(~ gsub("^payroll_", "", .), matches("^payroll_"))

mlb_top_teams <- mlb_yearly_increase %>%
  group_by(team) %>%
  summarise(
    payroll_1998 = first(`1998`),
    payroll_2014 = first(`2014`),
    pct_increase = ((payroll_2014 - payroll_1998) / payroll_1998) * 100
  ) %>%
  arrange(desc(pct_increase)) %>%
  head(3)

print(mlb_top_teams)

```

<h2>
<b> How are the metrics payroll_aggregate_computed and pct_increase reflected in the plot above, and how can
we see that the two sets of teams identified above are the top three in terms of these metrics?
</b>
</h2>
<h4> 
As mentioned in our initial plotting of payroll of each teams are distinct to each other as its horizontal line varied to each other, which implicates a different mean of payrolls.
<br>
The <b> Boston Red Sox, Los Angeles Dodgers, and New York Yankees </b> are identified using dplyr as the teams with the <b> highest payroll_aggregate_computed values</b>, characterized by high and varying payroll figures over the years.
<br>
While, The pct_increase metric shows the percentage increase in payroll from 1998 to 2014, with the <b>Washington Nationals, Detroit Tigers, and Philadelphia Phillies </b> showing <b>the greatest percentage increases </b>, indicating substantial growth in payroll over the analyzed period.


The plot shows how team payroll fluctuates over time, with teams with higher payroll aggregate values or significant pct_increase values easily identified. This visual representation complements quantitative analysis using dplyr, providing a more comprehensive understanding of payroll dynamics in Major League Baseball, as teams with wider spreads identified.<br>
Additionally, we have shown that the top 3 of high payroll over the years ( highes payroll_aggregate_computed) is different with top 3 of the greatest percentage increases of payroll, thus as assumed the an implication of different mean of payrolls.

</h4>

<br><br>

<h1> 2.2 Win percentage across years (15 points) </h1>

<h3>
<li>  Plot pct_wins as a function of year for each of the 30 teams, faceting the plot by team and adding a red dashed horizontal line for the average pct_wins across years of each team. </li>
 </h3>

```{r pct_wins_acc_year}

library(ggplot2)

pct_wins_acc_year <- mlb_yearly %>%
  pivot_longer(cols = starts_with("pct_wins_"), names_to = "year", values_to = "pct_wins") %>%
  mutate(year = as.numeric(str_extract(year, "\\d+"))) %>%
  ggplot(aes(x = year, y = pct_wins)) +
  geom_point() +
  geom_hline(aes(yintercept = mean(pct_wins, na.rm = TRUE)), linetype = "dashed", color = "red") +
  facet_wrap(~team, scales = "free_y") +
  labs(title = "Pct Wins Across Years for Each Team",
       x = "Year",
       y = "Pct Wins") +
  theme_minimal()

pct_wins_acc_year_title <- textGrob("Figure 1.3", gp = gpar(fontsize = 12, fontface = "bold"))
grid.arrange( pct_wins_acc_year,
  pct_wins_acc_year_title,
  heights = c(4, 0.5) 
)

```


<br><br>


<h3>
<li>  Using dplyr, identify the three teams with the greatest pct_wins_aggregate_computed and print a table of these teams along with pct_wins_aggregate_computed. </li>
 </h3>


```{r winper_4}
library(ggplot2)
thee_top_teams <- mlb_aggregate_joined %>%
  top_n(3, pct_wins_aggregate_computed) %>%
  select(team, pct_wins_aggregate_computed)
print(thee_top_teams)
```

<br><br>


<h3>
<li>  Using dplyr, identify the three teams with the most erratic pct_wins across years(as measured by the standard deviation, call it pct_wins_sd) and print a table of these teams along with pct_wins_sd. </li>
 </h3>


```{r winper_5}
top_erratic_teams <- mlb_yearly %>%
  gather(key = "year", value = "pct_wins", starts_with("pct_wins_x")) %>%
  group_by(team) %>%
  summarise(pct_wins_sd = sd(pct_wins, na.rm = TRUE)) %>%
  top_n(3, pct_wins_sd) %>%
  select(team, pct_wins_sd)

print(top_erratic_teams)

```

<br><br>

Re-PLOT of TOP TEAMS

```{r winper_6}
selected_teams <- c("Atlanta Braves", "Boston Red Sox", "New York Yankees","Detroit Tigers","Houston Astros", "Seattle Mariners")

selected_teams_plot <- mlb_yearly %>%
  pivot_longer(cols = starts_with("pct_wins_"), names_to = "year", values_to = "pct_wins") %>%
  mutate(year = as.numeric(str_extract(year, "\\d+"))) %>%
  filter(team %in% selected_teams) %>%
  ggplot(aes(x = year, y = pct_wins)) +
  geom_point() +
  geom_hline(aes(yintercept = mean(pct_wins, na.rm = TRUE)), linetype = "dashed", color = "red") +
  facet_wrap(~team, scales = "free_y") +
  labs(title = "Pct Wins Across Years for Selected Teams",
       x = "Year",
       y = "Pct Wins") +
  theme_minimal()

selected_teams_title <- textGrob("Plot of Top (pct_wins_sd) and (pct_wins_aggregate_computed)", gp = gpar(fontsize = 12, fontface = "bold"))

grid.arrange(
  selected_teams_plot,
  selected_teams_title,
  heights = c(4, 0.5)
)
```


<h2>
<b> How are the metrics pct_wins_aggregate_computed and pct_wins_sd reflected in the plot above, and how can
we see that the two sets of teams identified above are the top three in terms of these metrics?

</h2>
</b>

<h4> 
<br>
The <b>"Atlanta Braves", "Boston Red Sox", "New York Yankees"</b> are identified using dplyr as the teams with the <b> highest pct_wins_aggregate_computed values</b>, characterized by high and varying wins figures over the years.
<br>
While, The pct_wins_sd metric shows the standard deviation in percentage win from 1998 to 2014, with the <b>"Detroit Tigers","Houston Astros", "Seattle Mariners" </b> showing <b>the greatest standard deviation in percentage win </b>. 

As we plot the Top 3 teams of 2 different characteristic, we can see as shown that the plotting implicates a plotting <b> higher than 0.50 </p>. Thus, indeed we can see why the team top the rank.
</h4>
<br><br>

<h1> 2.3 Win percentage versus payroll (15 points) <h1>

<h3>
<li> Create a scatter plot of pct_wins versus payroll based on the aggregated data, labeling each point with the team name using geom_text_repel from the ggrepel package and adding the least squares line.
</li>
 </h3>

```{r winper_6_7}

library(ggplot2)
library(ggrepel)

mlb_aggregate <- mlb_raw_aggre %>%
  select(team, starts_with("the_payroll"), matches("^X.*\\.pct$")) %>%
  rename_with(~ "payroll_aggregate", starts_with("the_payroll")) %>%
  mutate(pct_wins_aggregate = rowMeans(select(., matches("^X.*\\.pct$")))) %>%
  select(team, starts_with("payroll_aggregate"), pct_wins_aggregate)

scatter_plot <- ggplot(mlb_aggregate, aes(x = payroll_aggregate, y = pct_wins_aggregate)) +
  geom_point() +
  geom_text_repel(aes(label = team), box.padding = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +  
  labs(title = "Scatter Plot of pct_wins_aggregate vs. payroll_aggregate",
       x = "Payroll Aggregate",
       y = "Pct Wins Aggregate") +
  theme_minimal()


scatter_acc_year_title <- textGrob("Figure 1.4", gp = gpar(fontsize = 12, fontface = "bold"))
grid.arrange( scatter_plot,
  scatter_acc_year_title,
  heights = c(4, 0.5) 
)


```

<h2>
<b> Isthe relationship between payroll and pct_wins positive or negative? Isthis what you would expect, and why?
</h2>
</b>
<h4> 
The relationship between payroll and pct_wins is posssitive as shown in the plot. <br> We can see that the plotting in Figure 1.4 that the New York Yankees, Boston Red Sox, and Los Angeles Dodgers continuously increases as Payroll and Pct_Win proportionally increases. Just like what we have on the Top 3 in our <b> payroll_aggregate_computed </b> which are also  New York Yankees, Boston Red Sox, and Los Angeles Dodgers. <br>
However, we cannot really say that there's a relationship between with the payroll and pct_wins only if we use more tool like a Simple Linear Model, etc., to see if there is a relationship that would make the data indeed proportionally into each other.

</h4>
<br><br>


<h1> 2.4 Team efficiency (5 points) </h1>

<h3>
<li> Using dplyr, identify the three teams with the greatest efficiency, and print a table of these teams
along with their efficiency, as well as their pct_wins_aggregate_computed and payroll_aggregate_computed.
</li>
 </h3>


```{r winper_7}
mlb_aggregate_computedz <- mlb_aggregate_joined %>%
  group_by(team) %>%
  summarise(
    payroll_aggregate_computed = sum(across(starts_with("payroll_aggregate_")), na.rm = TRUE), 
    pct_wins_aggregate_computed = mean(sum(across(starts_with("pct_wins_aggregate_")), na.rm = TRUE), na.rm = TRUE) / 17, 
    efficiency = pct_wins_aggregate_computed / payroll_aggregate_computed  
  ) 

top_efficiency_teams <- mlb_aggregate_computedz %>%
  top_n(3, wt = efficiency) %>%
  arrange(desc(efficiency))

print(top_efficiency_teams)


```

<h2>
<b> In what sense do these three teams appear efficient in the previous plot?
</h2>
</b>
<h4>
In accordance with our previous plot Figure 1.4, to say that a team is efficient, the quotient of pct_wins_aggregate_computed divided by payroll_aggregate_computed would be big. In other words, their payroll might not be big however, the percentage of them winning is great. <br>
As seen in Figure 1.4 again, the most obvious team in the plot is <b> Oakland Athletics </b> as the x-axis (Payroll) of team may be low, however the y-axis (Percentafe Wins) is high. In which can appear as well into our Top Three Teams which are Miami Marlins, Oakland Athletics, and lastly, Tampa Bay Rays.
</h4>

